Type Visualization for R

### 2020-06-21 paper/main.tex
paper: Add Julia's edits

@@

@@

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remain difficult. Important dimensions of
  code, like run-time type data, are inscrutable without appropriate tools.
  We built {\sc TypeVis} to explore and analyze run-time types in
  the R programming language by means of visualization.
  Our system focuses on understanding and comparing the type signatures of
  function calls across the R ecosystem.
  {\sc TypeVis} uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a gradual type system
  currently being developed for R.
}

@@

Programming languages commonly evolve by decree. The designer
decides that a new feature is necessary or that a past feature was
ill-conceived, and thus the language and its users move forward.
Sometimes these changes are the result of rigorous theory
or the product of community feedback. However, language design is
rarely informed by empirical data about how programs are actually
written in practice.

For a popular language, data collection is no challenge. Vast
quantities of publicly available code are hosted on open source
repositories like Github and language-specific package servers like npm.
Rather, the difficulty lies with analyzing and interpreting the data
at hand. Code is complex and highly structured. Simple
semantically-ignorant metrics like ``lines of code''
carry little significance.

To analyze one aspect of programming languages, we built {\sc TypeVis},
an interactive visualization of run-time types for the language R.
{\sc TypeVis} builds on a dataset of concrete execution traces
over the most widely used libraries in the R ecosystem.
Our visualization supports three broad tasks:
filtering the dataset down to a digestible subset,
understanding the types of function arguments and returns,
and comparing across different functions.
For filtering, {\sc TypeVis} uses the treemap idiom
to convey the relative scale of different packages.
Once a specific function is selected,
type signatures are represented as flows over types
where the width of the flow is proportional to frequency.
These flows can be readily compared within the same function,
or across different functions.

Tasks enabled by {\sc TypeVis} can assist language designers
during many phases of development.
For example, exploratory data analysis can weed out designs that
are incompatible with existing programs.
In particular, the visualization is built to answer questions
relevant to gradual type system development.
Supplemental material, including the data and source code of
{\sc TypeVis}, is freely available: \todo{Fill with archival link}

@@

R is a dynamically typed programming language widely used for
statistical computing. Unlike statically typed languages,
R programs do not contain type annotations
and never pass through a typechecker.
At run-time,
functions may called with values of arbitrary type,
and may return values of arbitrary type.\footnote{
  We use the word ``type'' to mean what others may call a ``type tag.''
  This is a run-time concept---we do not assume a static type system
  because there is none. Indeed, the goal is to help design one!
  A ``type signature'' in this context is one particular configuration
  of argument and return types a function is called with at run-time.
}

For statically typed languages, the type system is usually
an inseparable part of the language. Since they are designed
at the same time, the type system precedes any code written
in the language.
A gradual type system, by contrast,
permits the integration of static and dynamic typing.
Such a type system is capable of seamlessly mixing typed and untyped regions of code.
Often, a gradual type system is designed atop
an existing dynamically typed language.
Since many R programs already exist,
a successful gradual type system must be capable of
typechecking idiomatic R code.
{\sc TypeVis} aims to clarify how existing R programs
use types to assist in this task.
While this is our motivation, the visualization's use
is not necessarily limited to this application.
Anyone who has an interest in how the real R programs use types
may find it useful.

@@

The data consists of traces from 412 packages
containing over $760,000$ lines of R code
and $534,000$ lines of native code.
All libraries were obtained from CRAN,
the most popular R code repository.
Packages hosted on CRAN must satisfy certain criteria
to ensure they are high quality.
Additionally, the subset of 412 packages chosen here
all had code coverage above 65\%.

Each execution trace was recorded by running
the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of ``raw'' type signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
Our dataset consists of a table of traces
that have been processed to reduce some fields.
Traces were generated via a dynamic analysis
that instruments the R virtual machine with hooks
that intercept the required data at run-time.

While {\sc TypeVis} has only been run with
data from R execution traces, the architecture is not
limited to R.
Any dynamically typed language with the right run-time
hooks could generate data for the tool.
As long as execution trace information is adheres to
the same database schema, {\sc TypeVis} will
work properly without any modifications.

@@

We interviewed the primary researcher developing R's gradual type system
to inventory relevant domain-specific tasks.
Additionally, we sought feedback throughout the creation of {\sc TypeVis}
to ensure it was adequately satisfying requirements.
The domain-specific goals can be reduced to:

@@

\begin{figure*}
 \centering
 \includegraphics[width=\linewidth]{img/typevis.png}
 \caption{The main view of {\sc TypeVis}, with type flows on the left and the package hierarchy on the right. Here, the {\tt length.unit} and {\tt valid.unit} functions from the {\tt grid} package are both selected and displayed simultaneously.}
 \label{fig:typevis}
\end{figure*}

{\sc TypeVis} is an interactive web-based visualization
that permits real-time exploration,
even with large datasets.
The default dataset contains over a million rows
based on the analysis of almost five hundred
different R packages.
Figure~\ref{fig:typevis} displays
the heart of {\sc TypeVis}
in its initial configuration.
Most aspects of the system can be tweaked
by a user if desired.

@@

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the dataset was more deeply nested,
{\sc TypeVis} can be configured to
handle that.

If configured, {\sc TypeVis} allows for multiple function selection.
Both the {\tt length.unit} and {\tt valid.unit} functions
from the {\tt grid} package are selected in figure~\ref{fig:typevis}.
These functions are rendered simultaneously in the type flow
panel, allowing one to make comparisons between them.

@@

On the frontend, state management, interactivity,
and DOM manipulation is all done within the Vue framework.
This is in contrast with many web-based visualizations
that use D3 for DOM manipulation.
We chose this setup because of the many different
individual components whose state must be kept
synchronized.
This includes not only the visualizations, but also
user interface elements like the autocomplete search bar.
In our experience, Vue's model of reactivity is
more successful at maintaining consistency across the
entire system than using D3 and vanilla JavaScript.
While {\sc TypeVis} does not use D3 for DOM manipulation,
it still does use D3 for other visualization-related
computations.

@@

{\sc TypeVis} suffers from a number of shortcomings
and does not fully address the entire range of possible tasks
a language designer may need to perform.

@@

Filtering is an central mechanism in the system,
but it also comes at a cost.
If one has a specific function or several functions
in mind, then {\sc TypeVis} provides a useful
local view of that type information.
However, it does not support any kind of global
view of type information across a large amount of
functions. When the number of flows or nodes
becomes too great, the visualization will start
paginating, making comparisons over large amounts
of data impossible.
Aggregation and summarization would be key to
making a global view of the data feasible---that
remains future work.

@@

We have presented {\sc TypeVis}, an interactive visualization
of run-time type information for R.
It supports real-time exploration of a dataset of
execution traces from a large corpus of R packages.
To narrow the data under consideration,
{\sc TypeVis} provides facilities for filtering
and searching.
Type signatures are represented as flows traversing across
each argument type and return type in sequence.
Flow widths are proportional to how often
that type signature was seen during the dynamic analysis.
Interactions provide additional details and
assist users in comprehension and navigation.

Our tool is aimed at elucidating how R programmers
use the language in practice.
Specifically, the data and visualization will be used to
inform the design of a gradual type system for R.
However, {\sc TypeVis} is not tied to the R ecosystem
and is suitable to use for any dynamically typed language
with a similar dataset.
Additionally, the visualization could yield insights
useful for other purposes than gradual typing.

We hope that, in the future, programming language designers
will collect and use empirical data about how their languages
are used in practice.
Such data can be used to make informed decisions about
language design.
Visualization will be unavoidable in these scenarios
since semantically-meaning information about programs
is rich in structure.
Generic, canned visualizations will not be sufficient.
{\sc TypeVis} is just the first step in this direction.

\acknowledgments{
  The authors wish to thank the students of CS 7250 for useful design feedback,
  and contributors to all the open source projects that {\sc TypeVis} depends on.
  This work was supported in part by a Northeastern Graduate Fellowship.
}

>>>>

\usepackage{xspace}

@@

%% ===================================== Commands
\newcommand{\code}[1]{\texttt{\small{#1}}}
\newcommand{\typevis}{{\sc TypeVis}\xspace}

@@

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remain difficult. Important dimensions of
  code, like run-time type data, are inscrutable without appropriate tools.
  We built \typevis to explore and analyze run-time types in
  the R programming language by means of visualization.
  Our system focuses on understanding and comparing the type signatures of
  function calls across the R ecosystem.
  \typevis uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a gradual type system
  currently being developed for R.
}

@@

Programming languages commonly evolve by decree. Often, the designer
decides that a new feature is necessary or that a past feature was
ill-conceived. Thus, the language moves forward,
forcing its users to adapt to the changes.
Sometimes, these changes are the result of rigorous theory
or the product of community feedback. However, rarely is language design
informed by empirical data about how programs are \emph{actually
written} in practice.

%For a popular language, data collection is no challenge. Vast
Thanks to the large amount of open source code, collecting data about
the usage of a popular programming language is feasible. Vast
quantities of publicly available code are hosted on open source
repositories such as Github.
Furthermore, if there is a concern about the quality of such data
(for example, code duplication~\cite{lopes:2017}),
code can be obtained from curated
language-specific package repositories such as npm or CRAN.

To inform programming language design, all the collected data
needs to be analyzed and interpreted.
As code is complex and highly structured, simple semantically-ignorant
metrics like ``lines of code'' carry little significance.
Therefore, programming languages researchers often employ static
and dynamic analyses to gather information about specific aspects
of programs in the dataset, for example, the use of a language feature.
And even then it may be hard to make sense of the data,
especially if the dataset is large.
%Rather, the difficulty lies with analyzing and interpreting the data
%at hand. Code is complex and highly structured. Simple

To analyze one aspect of programming languages---\emph{the use of types}---we
built \typevis, an interactive visualization of run-time type
signatures of function calls.
%an interactive visualization of run-time types for the language R.
\typevis is instantiated for the language R:
it builds on a dataset of concrete execution traces over unit tests and vignettes
of the most widely used libraries in the R ecosystem.
For example, for the function call \code{!(FALSE)},
the recorded type signature would be \code{logical->logical},
meaning that function \code{!} took a logical value and returned a logical value.

The visualization supports three broad tasks:
\begin{itemize}
  \item filtering the dataset down to an amenable subset;
  \item understanding the types of function arguments and returns;
  \item comparing types across different functions.
\end{itemize}
%For filtering, \typevis uses the treemap idiom
%to convey the relative scale of different packages.
For exploring and filtering the dataset, \typevis relies on the treemap idiom
to represent R packages and functions, with the block sizes conveying
the relative popularity in the dataset.
In addition to the treemaps, a search box can be used to directly filter
by functions.
Once a specific subset of one or more functions is selected,
their type signatures are represented as flows over types
where the width of the flow is proportional to frequency.
These flows can be readily compared within the same function,
or across different functions.

Tasks enabled by \typevis can assist language designers
during multiple phases of development.
For example, exploratory data analysis can identify unexpected uses
of specific functions, or weed out language designs
that are incompatible with existing programs.
In particular, the visualization is built to answer questions
relevant to the development of a gradual type system for R.
However, the visualization is not limited to R and can be used to
analyze type usage in any programming language where type information
is available at run-time.
Supplemental material, including the data and source code of
\typevis, is freely available: \todo{Fill with archival link}

@@

R is a dynamically typed programming language widely used for
statistical computing. Unlike statically typed languages,
R programs do not contain type annotations
and never pass through a typechecker.
At run-time,
functions may be called with values of arbitrary type,
and may return values of arbitrary type.\footnote{
  We use the word ``type'' to mean what others may call a ``type tag.''
  This is a run-time concept---we do not assume a static type system
  because there is none. Indeed, the goal is to help design one!
  A ``type signature'' in this context is one particular configuration
  of argument and return types a function is called with at run-time.
}

Being a dynamically typed language most actively used in a single domain,
R is a good target for a gradual type system.
For statically typed languages, the type system is usually
an inseparable part of the language. Since they are designed
at the same time, the (static) type system precedes any code written
in the language.
A gradual type system, by contrast,
permits the integration of static and dynamic typing.
Such a type system is capable of seamlessly mixing typed and untyped regions of code.
Often, a gradual type system is designed atop
an existing dynamically typed language.
Since many R programs already exist,
a successful gradual type system must be capable of
typechecking idiomatic R code.
\typevis aims to clarify how existing R programs
use types to assist in this task.
While this is our motivation, the visualization's use
is not necessarily limited to this application.
Anyone who has an interest in how the real R programs use types
may find it useful.

@@

The data is provided by Alexi Turcotte:
it consists of traces from 412 packages
containing over $760,000$ lines of R code
and $534,000$ lines of native code.
All libraries were obtained from CRAN,
the most popular R code repository.
Packages hosted on CRAN must satisfy certain criteria
to ensure they are high quality.
Additionally, the subset of 412 packages chosen here
all had code coverage above 65\%.\todo{I don't understand what subset? Should we remove this line?}

Each execution trace was recorded by a dynamic analysis
running the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of ``raw'' function call signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
The type tags are assigned by the dynamic analysis and
refine over actual type tags in R; for example, the analysis
assigns scalar tag \code{double} to a singleton vector of double values
because R does not have native support for scalars.
Our dataset consists of a table of reduced execution traces:
because \typevis is designed for the analysis of types, not all information
in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

While \typevis has only been instantiated with
data from R execution traces, the architecture is not
limited to R.
Any dynamically typed language %with the right run-time
%hooks could generate data for the tool.
(or a statically typed language with run-time tag information)
could be instrumented to generate data for the tool.
As long as execution trace information adheres to
the same database schema, \typevis will
work properly without any modifications.

@@

We interviewed the primary researcher developing R's gradual type system
to inventory relevant domain-specific tasks.
Additionally, we sought feedback throughout the creation of \typevis
to ensure it was adequately satisfying requirements.
The domain-specific goals can be reduced to:

@@

\begin{figure*}
 \centering
 \includegraphics[width=\linewidth]{img/typevis.png}
 \caption{The main view of \typevis, with type flows on the left and the package hierarchy on the right. Here, the {\tt length.unit} and {\tt valid.unit} functions from the {\tt grid} package are both selected and displayed simultaneously.}
 \label{fig:typevis}
\end{figure*}

\typevis is an interactive web-based visualization
that permits real-time exploration,
even with large datasets.
The default dataset contains over a million rows
based on the analysis of almost five hundred
different R packages.
Figure~\ref{fig:typevis} displays
the heart of \typevis
in its initial configuration.
Most aspects of the system can be tweaked
by a user if desired.

@@

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the dataset was more deeply nested,
\typevis can be configured to
handle that.

If configured, \typevis allows for multiple function selection.
Both the {\tt length.unit} and {\tt valid.unit} functions
from the {\tt grid} package are selected in figure~\ref{fig:typevis}.
These functions are rendered simultaneously in the type flow
panel, allowing one to make comparisons between them.

@@

On the frontend, state management, interactivity,
and DOM manipulation is all done within the Vue framework.
This is in contrast with many web-based visualizations
that use D3 for DOM manipulation.
We chose this setup because of the many different
individual components whose state must be kept
synchronized.
This includes not only the visualizations, but also
user interface elements like the autocomplete search bar.
In our experience, Vue's model of reactivity is
more successful at maintaining consistency across the
entire system than using D3 and vanilla JavaScript.
While \typevis does not use D3 for DOM manipulation,
it still does use D3 for other visualization-related
computations.

@@

\typevis suffers from a number of shortcomings
and does not fully address the entire range of possible tasks
a language designer may need to perform.

@@

Filtering is an central mechanism in the system,
but it also comes at a cost.
If one has a specific function or several functions
in mind, then \typevis provides a useful
local view of that type information.
However, it does not support any kind of global
view of type information across a large amount of
functions. When the number of flows or nodes
becomes too great, the visualization will start
paginating, making comparisons over large amounts
of data impossible.
Aggregation and summarization would be key to
making a global view of the data feasible---that
remains future work.

@@

We have presented \typevis, an interactive visualization
of run-time type information for R.
It supports real-time exploration of a dataset of
execution traces from a large corpus of R packages.
To narrow the data under consideration,
\typevis provides facilities for filtering
and searching.
Type signatures are represented as flows traversing across
each argument type and return type in sequence.
Flow widths are proportional to how often
that type signature was seen during the dynamic analysis.
Interactions provide additional details and
assist users in comprehension and navigation.

Our tool is aimed at elucidating how R programmers
use the language in practice.
Specifically, the data and visualization will be used to
inform the design of a gradual type system for R.
However, \typevis is not tied to the R ecosystem
and is suitable to use for any dynamically typed language
with a similar dataset.
Additionally, the visualization could yield insights
useful for other purposes than gradual typing.

We hope that, in the future, programming language designers
will collect and use empirical data about how their languages
are used in practice.
Such data can be used to make informed decisions about
language design.
Visualization will be unavoidable in these scenarios
since semantically-meaning information about programs
is rich in structure.
Generic, canned visualizations will not be sufficient.
\typevis is just the first step in this direction.

\acknowledgments{
  The authors wish to thank the students of CS 7250 for useful design feedback,
  and contributors to all the open source projects that \typevis depends on.
  This work was supported in part by a Northeastern Graduate Fellowship.
}
